A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-24 MS.tex
Added top vs genus vs species elpd graph and changed a few estimates from medians to means

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.5
days (mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 9.5 days (mean estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.3 days (mean estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 7.0 days (mean estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 5.0 days (mean estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 12.1 days
(mean estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (mean estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

MRR estimates the length of time a mosquito remains alive and is still
in the area available for recapture. It thus provides a lower bound to mean
lifespan which we shall refer to as LBL. In 187 of the 230 MRR time
series the estimated LBL was less than 10 days (Fig. \ref{fig:mrr_lifespan_individualEstimates}). The smallest
estimate was 1.1 days for the Asian malaria vector
\emph{Anopheles subpictus s.l.} which is unfeasibly short and almost
certainly reflects dispersal out of the recapture zone or a violation of
the assumptions of our analyses. The longest estimate was 26.9 days for
the temperate species \emph{Aedes simpsoni s.l.} which is a vector of
yellow fever in Africa. There are multiple data sets for the most
important vector species such as \emph{Anopheles gambiae, Aedes aegypti}
and \emph{albopictus} and \emph{Culex tarsalis} all of which show
considerable variation. For example, there are 54 estimates of LBL for
\emph{Ae. aegypti} which range from 2.2 days to 38.3 days with a mean of
8.3 days and coefficient of variation of 0.7. There were significant
differences in LBL amongst species (ANOVA on median LBL controlling for
sex and pre-release feeding: \emph{F}\textsubscript{37,194} = 2.5,
\emph{p} \textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{38}\), \emph{p}\textless{}0.01).

The estimated mean LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.9, 6.8 and 8.1 days respectively with an overall estimate of 6.0 days
(Fig. \ref{fig:mrr_lifespans}; Table \ref{S-tab:mrr_estimated_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

The MRR experiments included a mixture of male-only and female-only
releases, and releases of both sexes. We estimated average male and
female LBL at the genus level (Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}; there were too few studies to
make comparisons at the species level). There was a consistent trend for
females to live longer than males for each of the genera, with the
difference largest for \emph{Aedes} (2.5 days; fraction of pairwise
posterior samples of females versus males where difference was less than
zero, p\textless{}0.01), followed by \emph{Anopheles} (2.0 days; p=0.17)
and \emph{Culex} (0.3 days; p=0.34). Overall, female mosquitoes were
estimated to live 0.9 days longer than males (p=0.10).

The MRR experiments included information on whether mosquitoes were
pre-fed with sugar (41 time series), blood (71), both (4) or
alternatively unfed (116). We estimate that female mosquitoes that were
fed on sugar pre-release lived on average for 0.7 days longer than those
that were not fed (Fig. \ref{S-fig:mrr_female_blood_sugar}; a pattern that was consistent across the
genera). There were insufficient males that were either fed or unfed
with sugar prior to release to make a meaningful comparison. Females
that were blood-fed prior to release on average lived 1.5 days longer
than those who were not fed for \emph{Aedes} but this trend was reversed
for \emph{Anopheles} meaning that there was little difference overall.

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). Table \ref{S-tab:dissection_estimated_lifespans_chron} provides posterior summaries of chronological for the species and genera in the dissection dataset (see also Fig. ). For ten species, we had enough data from both species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - there was a significant difference in the
time-series level LBLs (Table \ref{S-tab:comparison}).

@@

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with a posterior median LBL of 12.4 days (mean estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 11.9 days (mean
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara \citep{sinka2010dominant}, where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.5 days
(mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 9.5 days (mean estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in estimates was for
\emph{Aedes}, where the estimates from the MRR studies (8.1 days) are considerably longer than those of dissection-based studies (3.5 days).
This was followed by \emph{Culex} (a posterior mean of 2.9 days from the
MRR versus 4.9 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (6.8 versus 6.4 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 6.0 days versus 5.5
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) are similar or exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

@@

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{S-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 0.9 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquito ecology.

>>>>

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.5
days (mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 9.5 days (mean estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.3 days (mean estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 7.0 days (mean estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 5.0 days (mean estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 12.1 days
(mean estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
1.2 days on average (mean estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

MRR estimates the length of time a mosquito remains alive and is still
in the area available for recapture. It thus provides a lower bound to mean
lifespan which we shall refer to as LBL. In 187 of the 230 MRR time
series the estimated LBL was less than 10 days (Fig. \ref{fig:mrr_lifespan_individualEstimates}). The smallest
estimate was 1.1 days for the Asian malaria vector
\emph{Anopheles subpictus s.l.} which is unfeasibly short and almost
certainly reflects dispersal out of the recapture zone or a violation of
the assumptions of our analyses. The longest estimate was 26.9 days for
the temperate species \emph{Aedes simpsoni s.l.} which is a vector of
yellow fever in Africa. There are multiple data sets for the most
important vector species such as \emph{Anopheles gambiae, Aedes aegypti}
and \emph{albopictus} and \emph{Culex tarsalis} all of which show
considerable variation. For example, there are 54 estimates of LBL for
\emph{Ae. aegypti} which range from 2.5 days to 42.1 days with a mean of
11.4 days and coefficient of variation of 0.6 (all estimates are posterior mean). There were significant
differences in LBL amongst species (ANOVA on median LBL controlling for
sex and pre-release feeding: \emph{F}\textsubscript{37,194} = 2.5,
\emph{p} \textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{38}\), \emph{p}\textless{}0.01).

The estimated mean LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.9, 6.8 and 8.1 days respectively with an overall estimate of 6.0 days
(Fig. \ref{fig:mrr_lifespans}; Table \ref{S-tab:mrr_estimated_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
\ref{S-fig:mrr_genusTopLevel}; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

The MRR experiments included a mixture of male-only and female-only
releases, and releases of both sexes. We estimated average male and
female LBL at the genus level (Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}; there were too few studies to
make comparisons at the species level). There was a consistent trend for
females to live longer than males for each of the genera, with the
difference largest for \emph{Aedes} (2.9 days; fraction of pairwise
posterior samples of females versus males where difference was less than
zero, p\textless{}0.01), followed by \emph{Anopheles} (2.2 days; p=0.17)
and \emph{Culex} (0.2 days; p=0.34). Overall, female mosquitoes were
estimated to live 1.2 days longer than males (p=0.10).

The MRR experiments included information on whether mosquitoes were
pre-fed with sugar (41 time series), blood (71), both (4) or
alternatively unfed (116). We estimate that female mosquitoes that were
fed on sugar pre-release lived on average for 1.0 days (posterior mean) longer than those
that were not fed ($p>0.05$; Fig. \ref{S-fig:mrr_female_blood_sugar}; a pattern that was consistent across the
genera). There were insufficient males that were either fed or unfed
with sugar prior to release to make a meaningful comparison. Females
that were blood-fed prior to release on average lived 1.7 days (posterior mean) longer
than those who were not fed for \emph{Aedes} but this trend was reversed
for \emph{Anopheles} meaning that there was little difference overall (0.15 days; posterior mean; $p=0.44$).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). Table \ref{S-tab:dissection_estimated_lifespans_chron} provides posterior summaries of chronological for the species and genera in the dissection dataset (see also Fig. \ref{S-fig:dissection_lifetimes_exponential_chron}). For ten species, we had enough data from both species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - there was a significant difference in the
time-series level LBLs (Table \ref{S-tab:comparison}).

@@

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with an LBL of 12.4 days (mean estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 11.9 days (mean
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara \citep{sinka2010dominant}, where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.5 days
(mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 9.5 days (mean estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in estimates was for
\emph{Aedes}, where the estimates from the MRR studies (8.1 days) are considerably longer than those of dissection-based studies (3.5 days).
This was followed by \emph{Culex} (a posterior mean of 2.9 days from the
MRR versus 4.9 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (6.8 versus 6.4 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 6.0 days versus 5.5
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) are similar or exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

@@

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{S-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 1.2 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquito ecology.

### 2018-11-24 supplementary_materials.tex
Added top vs genus vs species elpd graph and changed a few estimates from medians to means

>>>>

\begin{figure}[h]
	\centerline{\includegraphics[width=1\textwidth]{./Figure_files/mrr_genus_vs_species_vs_top.pdf}}
	\caption{\textbf{The estimated predictive accuracy of the species- and overall-level models relative to the genus-level model.} The predictive accuracy was determined using K-Fold cross-validation as described in SOM; the measure of accuracy presented here by the central dots is the difference in estimated expected log pointwise predictive density compared to the exponential model. The lower and upper whiskers represent the lower and upper bounds of the 95\% confidence interval in predictive accuracy.}
	\label{fig:mrr_genusTopLevel}
\end{figure}
